---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true



---




```{r}
library(ggplot2)
library(scales)
library(Hmisc)  
```



---

#### Loading and preprocessing the data


First check to make sure file isn't already unzipped.  If it isn't, then unzip.  
If it is, read it in.


```{r}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}

dat = read.csv('activity.csv', header = T)
```




#### Processing / transforming the data into something usable for our purposes.



```{r}
#dat$interval <- strptime(gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", dat$interval), format='%H:%M')
```



#### What is mean total number of steps taken per day?



```{r}
stepsByDay <- tapply(dat$steps, dat$date, sum, na.rm=TRUE)
```



#### 1. Make a Histogram of the total number of steps taken each day



```{r}
qplot(stepsByDay, xlab='Total steps per day', ylab='Frequency using binwith 500', binwidth=750)
```



#### 2. Calculate and report the mean and median total number of steps taken per day.



```{r}
stepsByDayMean <- mean(stepsByDay)
stepsByDayMedian <- median(stepsByDay)
```

* Mean: `r stepsByDayMean`
* Median:  `r stepsByDayMedian`


---


#### What is the average daily activity pattern?




```{r}
averageStepsPerInt <- aggregate(x=list(meanSteps=dat$steps), by=list(interval=dat$interval), FUN=mean, na.rm=TRUE)
```

#### 1. Make a time series plot



```{r}
ggplot(data=averageStepsPerInt, aes(x=interval, y=meanSteps)) + geom_line() + 
        xlab("5-minute interval") + ylab("average number of steps taken") 
```


#### 2. Which 5-minute interval, on average across all the daays in the dataset, contains the 
#### maximum number of steps?



```{r}
maxSteps <- which.max(averageStepsPerInt$meanSteps)
timeMaxSteps <-  gsub("([0-9]{1,2})([0-9]{2})", 
                      "\\1:\\2",averageStepsPerInt[maxSteps,'interval'])
```

* Max steps are at: `r timeMaxSteps`



---



### Imputing missing values


#### 1. Calculate and report the total number of missing values in the dataset.

```{r}
totMissingValues <- length(which(is.na(dat$steps)))
```

* Total number of missing values: 'r totMissingValues'




#### 2. Devise a strategy for filling in all of the missing values in the dataset.

```{r}
datImputed <- dat
datImputed$steps <- impute(dat$steps, fun=mean)
```



#### 4. Make a histogram of the total number of steps taken each day and Calculate
####        and report the mean and median total number of steps taken per day.
        
```{r}
stepsByDayImputed <- tapply(datImputed$steps, datImputed$date, sum)
qplot(stepsByDayImputed, xlab='Total steps per day (Imputed)',
      ylab='Frequency using binwith 500', binwidth=500)

stepsByDayMeanImputed <- mean(stepsByDayImputed)
stepsByDayMedianImputed <- median(stepsByDayImputed)
```



##### Reporting Mean and Median for the new steps per day data.

* Mean (Imputed): `r stepsByDayMeanImputed`
* Median (Imputed):  `r stepsByDayMedianImputed`


####Do these values differ from the estimates from the first part of the assignment? 

Yes, they differ from the original values quite significantly.

#### What is the impact of imputing missing data on the estimates of the daily number of steps?
Filing in the missing data, increasing both the mean and median value.         



---



#### Are there differences in activity patterns between weekdays and weekends?



#### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
datImputed$dateType <-  ifelse(as.POSIXlt(datImputed$date)$wday %in% c(0,6), 'weekend', 'weekday')
```




#### 2. Make a panel plot containing a time series plot.

```{r}
averagedDatImputed <- aggregate(steps ~ interval + dateType, data=datImputed,mean)
ggplot(averagedDatImputed, aes(interval, steps)) + geom_line() + 
        facet_grid(dateType ~ .) +
        xlab("5-minute interval") +  ylab("avarage number of steps")
```